---
title: "Chapter 4: CV for map accuracy estimation"
output: html_document
date: "`r Sys.Date()`"
author: Carles Milà
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F, fig.align = 'center')
```

**Notes**: 

Shared legend at the cost of importing another package? cowplot or ggpubr.

Change to colour-blind safe palettes

### Setup and data

```{r libraries}
library(terra)
library(sf)
library(raster)
library(ggplot2)
library(gridExtra)
library(caret)
library(ranger)
library(knitr)
library(CAST)
```

```{r readdata}
# Read data
pts_clustered <- st_read("data/pts_clustered.gpkg", quiet = T)
pts_random <- st_read("data/pts_random.gpkg", quiet = T)
pts_validation <- st_read("data/pts_validation.gpkg", quiet = T)
aoi <- st_read("data/AOI.gpkg", quiet = T)
predictors <- rast("data/predictors.tif")
response <- rast("data/response.tif")

# Terra to CRS - plot_geodist doesn't support terra, needs to be fixed
# Also proj4 warning!?
predictors_r <- stack(predictors)
raster::crs(predictors_r) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0" 
response_r <- raster(response)
raster::crs(response_r) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0" 
```

### 4.0 Motivation

Motivate the problem by showing the difference in NND functions and ideally a random sample of validation points, which is not available in most cases.

In geographical space:

```{r 4.0 geo, fig.width=10, fig.height=5}
plt_rand_geo <- plot_geodist(pts_random, aoi, testdata = pts_validation, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom") + labs(fill = "") +
  ggtitle("Random points")
plt_clust_geo <- plot_geodist(pts_clustered, testdata = pts_validation, aoi, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom")  + labs(fill = "") +
  ggtitle("Clustered points")
grid.arrange(plt_rand_geo, plt_clust_geo, nrow = 1)
```

In feature space:

```{r 4.0 feature, fig.width=10, fig.height=5}
plt_rand_feat <- plot_geodist(pts_random, predictors_r, type = "feature",
                              testdata = pts_validation, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom") + scale_fill_discrete(name = "") +
  ggtitle("Random points")
plt_clust_feat <- plot_geodist(pts_clustered, predictors_r, type = "feature",
                               testdata = pts_validation, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom") + scale_fill_discrete(name = "") +
  ggtitle("Clustered points")
grid.arrange(plt_rand_feat, plt_clust_feat, nrow = 1)
```

### 4.1 Random and spatial CV

Define 5 random folds:

```{r 4.1 randomfolds, fig.width=10, fig.height=5}
# Fold generation
set.seed(123)
pts_random$random_fold <- sample(rep(1:5, ceiling(nrow(pts_random)/5)),
                                 nrow(pts_random))
pts_random$random_fold <- as.factor(pts_random$random_fold)
pts_clustered$random_fold <- sample(rep(1:5, ceiling(nrow(pts_random)/5)),
                                    nrow(pts_random))
pts_clustered$random_fold <- as.factor(pts_clustered$random_fold)

# Cartographic representation
map1_random <- ggplot() +
  geom_sf(data = aoi) +
  geom_sf(data = pts_random, aes(col = random_fold)) +
  scale_color_discrete() +
  theme_bw() + theme(legend.position = "bottom")
map1_clustered <- ggplot() +
  geom_sf(data = aoi) +
  geom_sf(data = pts_clustered, aes(col = random_fold)) +
  theme_bw() + theme(legend.position = "bottom")
grid.arrange(map1_random, map1_clustered, nrow = 1)
```

Define 5 spatial folds:

```{r 4.1 spatialfolds, fig.width=10, fig.height=5}
# Fold generation
naive_spblocks <- function(tpoints, aoipoly, blocksize, k, seed){
  set.seed(seed)
  polygrid <- st_sf(geom = st_make_grid(aoipoly, blocksize))
  polygrid <- polygrid[apply(st_intersects(polygrid, tpoints, sparse = F), 1, any),]
  polygrid$spatial_fold <- sample(rep(1:k, ceiling(nrow(polygrid)/k)), nrow(polygrid))
  st_drop_geometry(st_join(tpoints, polygrid))$spatial_fold
}
pts_random$spatial_fold <- naive_spblocks(pts_random, aoi, 5, 5, 1234)
pts_random$spatial_fold <- as.factor(pts_random$spatial_fold)
pts_clustered$spatial_fold <- naive_spblocks(pts_clustered, aoi, 5, 5, 1234)
pts_clustered$spatial_fold <- as.factor(pts_clustered$spatial_fold)

# Cartographic representation
map2_random <- ggplot() +
  geom_sf(data = aoi) +
  geom_sf(data = pts_random, aes(col = spatial_fold)) +
  scale_color_discrete() +
  theme_bw() + theme(legend.position = "bottom")
map2_clustered <- ggplot() +
  geom_sf(data = aoi) +
  geom_sf(data = pts_clustered, aes(col = spatial_fold)) +
  theme_bw() + theme(legend.position = "bottom")
grid.arrange(map2_random, map2_clustered, nrow = 1)
```

Create CV lists by using CAST:

```{r CASTfunction}
rand_rCV_folds <- CreateSpacetimeFolds(st_drop_geometry(pts_random), spacevar = "random_fold", k = 5)
clust_rCV_folds <- CreateSpacetimeFolds(st_drop_geometry(pts_clustered), spacevar = "random_fold", k = 5)
rand_sCV_folds <- CreateSpacetimeFolds(st_drop_geometry(pts_random), spacevar = "spatial_fold", k = 5) 
clust_sCV_folds <- CreateSpacetimeFolds(st_drop_geometry(pts_clustered), spacevar = "spatial_fold", k = 5)
```

In geographical space:

```{r 4.1 geo, fig.width=10, fig.height=8}
plt_rand_geo1 <- plot_geodist(pts_random, aoi, 
                              cvfolds = rand_rCV_folds$indexOut, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom") + labs(fill = "") +
  ggtitle("Random points, random CV")
plt_clust_geo1 <- plot_geodist(pts_clustered, aoi,
                               cvfolds = clust_rCV_folds$indexOut, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom")  + labs(fill = "") +
  ggtitle("Clustered points, random CV")
plt_rand_geo2 <- plot_geodist(pts_random, aoi, 
                              cvfolds = rand_sCV_folds$indexOut, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom") + labs(fill = "") +
  ggtitle("Random points, spatial CV")
plt_clust_geo2 <- plot_geodist(pts_clustered, aoi, 
                               cvfolds = clust_sCV_folds$indexOut, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom")  + labs(fill = "") +
  ggtitle("Clustered points, spatial CV")
grid.arrange(plt_rand_geo1, plt_clust_geo1, plt_rand_geo2, plt_clust_geo2, nrow = 2)
```

In feature space:

```{r 4.1 feature, fig.width=10, fig.height=8}
plt_rand_feat1 <- plot_geodist(pts_random, predictors_r, type = "feature",
                               cvfolds = rand_rCV_folds$indexOut, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom") + labs(fill = "") +
  ggtitle("Random points, random CV")
plt_clust_feat1 <- plot_geodist(pts_clustered, predictors_r, type = "feature",
                                cvfolds = clust_rCV_folds$indexOut, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom")  + labs(fill = "") +
  ggtitle("Clustered points, random CV")
plt_rand_feat2 <- plot_geodist(pts_random, predictors_r, type = "feature",
                               cvfolds = rand_sCV_folds$indexOut, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom") + labs(fill = "") +
  ggtitle("Random points, spatial CV")
plt_clust_feat2 <- plot_geodist(pts_clustered, predictors_r, type = "feature", 
                                cvfolds = clust_sCV_folds$indexOut, showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom")  + labs(fill = "") +
  ggtitle("Clustered points, spatial CV")
grid.arrange(plt_rand_feat1, plt_clust_feat1, plt_rand_feat2, plt_clust_feat2, nrow = 2)
```

### 4.2 Nearest neighbour distance matching cross-validation

```{r nndm}
rand_nndm_folds <- nndm(pts_random, aoi, min_train = 0.6)
clust_nndm_folds <- nndm(pts_clustered, aoi, min_train = 0.6)
```

In geographical space:

```{r 4.2 geo, fig.width=10, fig.height=4}
plt_rand_geo1 <- plot_geodist(pts_random, aoi, 
                              cvfolds = rand_nndm_folds$indx_test, 
                              cvtrain = rand_nndm_folds$indx_train,
                              showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom") + labs(fill = "") +
  ggtitle("Random points, NNDM LOO CV")
plt_clust_geo1 <- plot_geodist(pts_clustered, aoi,
                               cvfolds = clust_nndm_folds$indx_test, 
                               cvtrain = clust_nndm_folds$indx_train,
                               showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom")  + labs(fill = "") +
  ggtitle("Clustered points, NNDM LOO CV")
grid.arrange(plt_rand_geo1, plt_clust_geo1, nrow = 1)
```

In feature space:

```{r 4.2 feature, fig.width=10, fig.height=4}
plt_rand_feat1 <- plot_geodist(pts_random, predictors_r, type = "feature",
                               cvfolds = rand_nndm_folds$indx_test, 
                               cvtrain = rand_nndm_folds$indx_train,
                               showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom") + labs(fill = "") +
  ggtitle("Random points, NNDM LOO CV")
plt_clust_feat1 <- plot_geodist(pts_clustered, predictors_r, type = "feature",
                                cvfolds = clust_nndm_folds$indx_test, 
                                cvtrain = clust_nndm_folds$indx_train,
                                showPlot = F)$plot +
  theme_bw() +
  theme(legend.position = "bottom")  + labs(fill = "") +
  ggtitle("Clustered points, NNDM LOO CV")
grid.arrange(plt_rand_feat1, plt_clust_feat1, nrow = 1)
```

### 4.3 Model fitting and validation

Common objects:

```{r commonparams}
pgrid <- data.frame(mtry = 5, min.node.size = 5, splitrule = "variance")
form <- paste0("suitability ~ " , paste0("bio", c("1", "2", "5", "6", "7", "10",
                                                 "11", "12", "13", "14", "15", "18", 
                                                 "19"), collapse = " + "),
               " + lon + lat")
form <- as.formula(form)
response_df <- as.data.frame(c(response, predictors))
response_df <- response_df[apply(response_df, 1, function(x) all(!is.nan(x))),]
```

Model 1: Random samples

```{r modrand}
cntrl_mod <- trainControl(method = "cv", number = 10, allowParallel = TRUE)
rand_mod <- train(form, data = st_drop_geometry(pts_random), method="ranger", 
                  trControl = cntrl_mod, tuneGrid = pgrid)
```

Evaluation 1.1: true map accuracy

```{r ev1.1}
rand_true <- predict(rand_mod, newdata = response_df)
rand_true_rmse <- sqrt(mean((response_df$suitability - rand_true)^2))
```

Evaluation 1.2: probability sample

```{r ev1.2}
rand_prob <- predict(rand_mod, newdata = pts_validation)
rand_prob_rmse <- sqrt(mean((pts_validation$suitability - rand_prob)^2))
```

Evaluation 1.3: random CV

```{r ev1.3}
cntrl_mod <- trainControl(method = "cv", index = rand_rCV_folds$index, allowParallel = TRUE,
                          savePredictions='final')
rand_rCV <- train(form, data = st_drop_geometry(pts_random), method="ranger", 
                  trControl = cntrl_mod, tuneGrid = pgrid)
rand_rCV_rmse <- global_validation(rand_rCV)["RMSE"]
```

Evaluation 1.4: spatial CV

```{r ev1.4}
cntrl_mod <- trainControl(method = "cv", index = rand_sCV_folds$index, allowParallel = TRUE,
                          savePredictions='final')
rand_sCV <- train(form, data = st_drop_geometry(pts_random), method="ranger", 
                  trControl = cntrl_mod, tuneGrid = pgrid)
rand_sCV_rmse <- global_validation(rand_sCV)["RMSE"]
```

Evaluation 1.5: NNDM CV

```{r ev1.5}
cntrl_mod <- trainControl(method = "cv", 
                          index = rand_nndm_folds$indx_train,
                          indexOut = rand_nndm_folds$indx_test,
                          allowParallel = TRUE, savePredictions='final')
rand_nndmCV <- train(form, data = st_drop_geometry(pts_random), method="ranger", 
                     trControl = cntrl_mod, tuneGrid = pgrid)
rand_nndmCV_rmse <- global_validation(rand_nndmCV)["RMSE"]
```

Model 2: Clustered samples

```{r modclust}
cntrl_mod <- trainControl(method = "cv", number = 10, allowParallel = TRUE)
clust_mod <- train(form, data = st_drop_geometry(pts_clustered), method="ranger", 
                   trControl = cntrl_mod, tuneGrid = pgrid)
```

Evaluation 2.1: true map accuracy

```{r ev2.1}
clust_true <- predict(clust_mod, newdata = response_df)
clust_true_rmse <- sqrt(mean((response_df$suitability - clust_true)^2))
```

Evaluation 2.2: probability sample

```{r ev2.2}
clust_prob <- predict(clust_mod, newdata = pts_validation)
clust_prob_rmse <- sqrt(mean((pts_validation$suitability - clust_prob)^2))
```

Evaluation 2.3: random CV

```{r ev2.3}
cntrl_mod <- trainControl(method = "cv", index = clust_rCV_folds$index, allowParallel = TRUE,
                          savePredictions='final')
clust_rCV <- train(form, data = st_drop_geometry(pts_clustered), method="ranger", 
                  trControl = cntrl_mod, tuneGrid = pgrid)
clust_rCV_rmse <- global_validation(clust_rCV)["RMSE"]
```

Evaluation 2.4: spatial CV

```{r ev2.4}
cntrl_mod <- trainControl(method = "cv", index = clust_sCV_folds$index, allowParallel = TRUE,
                          savePredictions='final')
clust_sCV <- train(form, data = st_drop_geometry(pts_clustered), method="ranger", 
                  trControl = cntrl_mod, tuneGrid = pgrid)
clust_sCV_rmse <- global_validation(clust_sCV)["RMSE"]
```

Evaluation 2.5: NNDM CV

```{r ev2.5}
cntrl_mod <- trainControl(method = "cv", 
                          index = clust_nndm_folds$indx_train,
                          indexOut = clust_nndm_folds$indx_test,
                          allowParallel = TRUE, savePredictions='final')
clust_nndmCV <- train(form, data = st_drop_geometry(pts_clustered), method="ranger", 
                      trControl = cntrl_mod, tuneGrid = pgrid)
clust_nndmCV_rmse <- global_validation(clust_nndmCV)["RMSE"]
```

Table of results RMSE:

```{r tabres}
resmod <- data.frame(Sampling = c("Random", "Clustered"),
                     True = c(rand_true_rmse, clust_true_rmse),
                     ProbSample = c(rand_prob_rmse, clust_prob_rmse),
                     rCV = c(rand_rCV_rmse, clust_rCV_rmse),
                     sCV = c(rand_sCV_rmse, clust_sCV_rmse),
                     NNDM = c(rand_nndmCV_rmse, clust_nndmCV_rmse))
kable(resmod, digits = 3)
```